---
title: "Exploratory analysis"
author: "Esther Kyeremah"
date: "2023-11-01"
output: github_document
---


```{r}
library(tidyverse)
library(ggridges)
```

##Setting options

```{r, eval = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  gif.asp = .6,
  out.width = "90%"
)


theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.comtinuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

## Load weather data
 
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
month = lubridate::floor_date(date, unit = "month")) %>% 
  select(name, id, everything())
```


##Group BY
```{r}
weather_df %>% 
  group_by(name, month)
```
So we see that theyve written groups: name [3] under the tibble(easy to miss).

you can ungroup after doing whatever analysis too.. and you can ungroup for just one or more variables

```{r}
weather_df %>% 
group_by(name, month) %>% 
ungroup(name)
```


##counting things

count name or month or both name and month observations
```{r}
weather_df %>% 
group_by(name) %>% 
  summarize(n_obs = n())

weather_df %>% 
group_by(month) %>% 
  summarize(n_obsi = n())

weather_df %>% 
group_by(name, month) %>% 
  summarize(n_obsis = n())

```



we can just use the count function

```{r}
weather_df %>% 
  count(name, month, name = "n_obs")
```

other helpful counters
```{r}
weather_df %>% 
group_by(month) %>% 
  summarize(
    n_obs = n(),
    n_days = n_distinct(date))
```


##2by2 tables
```{r}
weather_df %>% 
  filter(name != "Molokai_HI") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold",
      TRUE ~ ""
      )
  ) %>% 
 janitor::tabyl(name, cold)
  
```


General summaries

```{r}
weather_df %>% 
group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax)
  )
``` 

there are missing vslues so you can get rid of them

```{r}
weather_df %>% 
group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  )
``` 

THIS IS A DATAFRAME!
```{r}
weather_df %>% 
group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  )

``` 

We can make a plot from the summary
```{r}
weather_df %>% 
group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  ) %>% 
ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()  
```


what is below shows what we would have had if we 
had just put in our raw date and tmax..it literally gives us the the data day by day so depending on what you want to convey in your research
```{r}
weather_df %>% 
ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line()  
```

 Suppose you want to summarize many columns
```{r}
 weather_df %>% 
group_by(name, month) %>% 
summarize(across(prcp:tmin, mean))
```

Reminder: sometimes your dresults are easier to read in another format
```{r}
 weather_df %>% 
group_by(name, month) %>% 
summarize(mean_tmax = mean(tmax)) %>% 
  pivot_wider(
names_from = name,
values_from = mean_tmax) %>% 
knitr::kable(digits = 1)
```


